---
title: "Chapter 4 Non-Numeric Values"
output: html_notebook
---

# 4.1 Logical Values

Logical values -\> Boolean -\> `TRUE` or `FASLE`

## 4.1.1 `TRUE` or `FALSE`?

Can assign logical values like numerics

```{r}
foo <- TRUE
foo
```

Can use `F` or `T` instead

```{r}
bar <- F
bar
```

Can also fill arrays

```{r}
baz <- c(T, F, F, F, T, F, T, T, T, F, T, F)
baz
```

```{r}
length(x=baz)
```

```{r}
qux <- matrix(data=baz, nrow=3, ncol=4, byrow=foo)
qux
```

## 4.1.2 A Logical Outcome: Relational Operators

Typical relational operators:

-   `==`

-   `!=`

-   `>`

-   `<`

-   `>=`

-   `<=`

```{r}
1 == 2
```

```{r}
1 > 2
```

```{r}
(2 - 1) <= 2
```

```{r}
1 != (2 + 3)
```

Can be used with arrays as well

```{r}
foo <- c(3, 2, 1, 4, 1, 2, 1, -1, 0, 3)
bar <- c(4, 1, 2, 1, 1, 0, 0, 3, 0, 4)
length(x=foo) == length(x=bar)
```

Relational operators work element by element

```{r}
foo == bar
```

```{r}
foo < bar
```

```{r}
foo <= bar
```

```{r}
foo <= (bar + 10)
```

Vector recycling also applies

```{r}
baz <- foo[c(10, 3)]
baz
```

```{r}
foo > baz
```

```{r}
foo < 3
```

```{r}
foo.mat <- matrix(foo, nrow=5, ncol=2)
foo.mat
```

```{r}
bar.mat <- matrix(bar, nrow=5, ncol=2)
bar.mat
```

```{r}
foo.mat <= bar.mat
```

```{r}
foo.mat < 3
```

Can use `all()` to return if all of a vector is `TRUE` on not, `any()` to return if any are `TRUE`

```{r}
qux <- foo == bar
qux
```

```{r}
any(qux)
```

```{r}
all(qux)
```

```{r}
quux <- foo <= (bar + 10)
quux
```

```{r}
any(quux)
```

```{r}
all(quux)
```

## 4.1 Exercises

### (a) Store `c(6, 9, 7, 3, 6, 7, 9, 6, 3, 6, 6, 7, 1, 9, 1)`, then identify:

```{r}
A <- c(6, 9, 7, 3, 6, 7, 9, 6, 3, 6, 6, 7, 1, 9, 1)
```

`which` returns the indices that are `TRUE`

#### (i) Those equal to $6$

```{r}
which(A == 6)
```

#### (ii) Those greater than or equal to $6$

```{r}
which(A >= 6)
```

#### (iii) Those less than $6 + 2$

```{r}
which(A < (6 + 2))
```

#### (iv) Those not equal to $6$

```{r}
which(A != 6)
```

### (b) Create a new vector from (a) by deleting the first three elements. Fill a $2 \times 2 \times 3$ array. Then identify:

```{r}
B <- array(data=A[-1:-3], dim=c(2, 2, 3))
```

#### (i) Those less than or equal to $6$ divided by $2$, plus $4$

```{r}
B[B <= ((6 / 2) + 4)]
```

#### (ii) Those less than or equal to $6$ divided by $2$, plus $4$, after increasing every element by $2$

```{r}
B[B + 2 <= ((6 / 2) + 4)]
```

### (c) Check the specific locations of elements equal to 0 in $I_{10}$

```{r}
diag(10) == 0
```

### (d) Check whether any of the values of the logical arrays created in (b) are `TRUE`, if they are check if they are all `TRUE`

```{r}
print(any(B <= ((6 / 2) + 4)))
print(all(B <= ((6 / 2) + 4)))

print(any(B + 2 <= ((6 / 2) + 4)))
print(all(B + 2 <= ((6 / 2) + 4)))
```

### (e) By extracting the diagonal elements of the logical matrix in (c), use `any()` to check there are no `TRUE` entries

```{r}
any(diag(diag(10) == 0))
```

## 4.1.3 Multiple Comparisons: Logical Operators

Can compare element-wise or as a single comparison:

-   `&` element-wise

-   `&&` single

-   `|` element-wise

-   `||` single

-   `!`

Single comparitiors on scalars

```{r}
FALSE || ((T && TRUE) || FALSE)
```

```{r}
!TRUE && TRUE
```

```{r}
(T && (TRUE || F)) && FALSE
```

```{r}
(6 < 4) || (3 != 1)
```

Element-wise on arrays

```{r}
foo <- c(T, F, F, F, T, F, T, T, T, F, T, F)
foo
```

```{r}
bar <- c(F, T, F, T, F, F, F, F, T, T, T, T)
bar
```

```{r}
foo & bar
```

```{r}
foo | bar
```

Single comparators on arrays, only the first elements

```{r}
foo && bar
```

```{r}
foo || bar
```

## 4.2 Exercises

### (a) Store `c(7, 1, 7, 10, 5, 9, 10, 3, 10, 8)`. Identify elements greater than $5$ OR equal to $2$

```{r}
A <- c(7, 1, 7, 10, 5, 9, 10, 3, 10, 8)
which((A > 5) | (A == 2))
```

### (b) Store `c(8, 8, 4, 4, 5, 1, 5, 6, 6, 8)`. Identify elements less than or equal to $6$ AND not equal to $4$

```{r}
B <- c(8, 8, 4, 4, 5, 1, 5, 6, 6, 8)
which((B <= 6) & (B != 4))
```

### (c) Identify the elements that satisfy (a) AND (b)

```{r}
C_1 <- (A > 5) | (A == 2)
C_2 <- (B <= 6) & (B != 4)
C_1 & C_2
```

### (d) Store a third vector equal to the element-wise sum of (a) and (b), then determine:

```{r}
D <- A + B
```

#### (i) The elements greater than or equal to $14$ but not equal to $15$

```{r}
which((D >= 14) & (D != 15))
```

#### (ii) The elements of the element-wise division of (d) by (a) that are greater than or equal to $4$ OR less than or equal to $2$

```{r}
which(((D / A) > 4) | ((D / A) <= 2))
```

### (e) Confirm using the long version in all of the above, only performs the first comparison

```{r}
print(((A > 5) || (A == 2)) == 
          ((A[1] > 5) || (A[1] == 2)))

print(((B <= 6) && (B != 4)) ==
          ((B[1] <= 6) && (B[1] != 4)))

print((C_1 && C_2) ==
          (C_1[1] && C_2[1]))

print(((D >= 14) && (D != 15)) ==
          ((D[1] >= 14) && (D[1] != 15)))

print((((D / A) > 4) || ((D / A) <= 2)) ==
          (((D[1] / A[1]) > 4) || ((D[1] / A[1]) <= 2)))

```

## 4.1.4 Logicals Are Numbers

`TRUE` = $1$

`FALSE` = $0$

```{r}
TRUE + TRUE
```

```{r}
FALSE - TRUE
```

```{r}
T + T + F + T + F + F + T
```

Can interchange with numerical values

```{r}
1 && 1
```

```{r}
1 || 0
```

```{r}
0 && 1
```

## 4.1.5 Logical Sub-setting and Extraction

Can use logical values to subset a vector

```{r}
my_vec <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)
my_vec[c( F, T, F, F, F, F, F, F, F, T)]
```

Can use conditionals within the subset

```{r}
my_vec < 0
```

```{r}
my_vec[my_vec < 0]
```

Vector recycling also applies

```{r}
my_vec[c(T, F)]
```

```{r}
my_vec[(my_vec > 0) & (my_vec < 1000)]
```

Can be used to assign elements

```{r}
my_vec[my_vec < 0] <- -200
my_vec
```

Can use which to return `TRUE` indices

```{r}
which(x=c(T, F, F, T, T))
```

```{r}
which(x=(my_vec < 0))
```

Can then use to slice vector

```{r}
my_vec[-which(x=(my_vec < 0))]
```

Can use them to also slice arrays

```{r}
A <- matrix(
        c(0.3, 4.5, 55.3, 91, 0.1, 105.5, -4.2, 8.2, 27.9), 
        nrow=3, 
        ncol=3
        )
A
```

```{r}
A[c(T, F, F), c(F, T, T)]
```

```{r}
A < 1
```

```{r}
A[A < 1] <- -7
A
```

```{r}
A > 25
```

`which` defaults to treating arrays like vectors

```{r}
which(x=(A > 25))
```

```{r}
which(x=(A > 25), arr.ind=T)
```

## 4.3 Exercises

### (a) Store `c(7, 5, 6, 1, 2, 10, 8, 3, 8, 2)`, then do:

```{r}
A <- c(7, 5, 6, 1, 2, 10, 8, 3, 8, 2)
```

#### (i) Extract elements greater than or equal to $5$, store it.

```{r}
A_1 <- A[A >= 5]
A_1
```

#### (ii) Display elements that remain after removing all elements greater than or equal to $5$

```{r}
A[-which(A >= 5)]
```

### (b) Use (a)(i) to make a $2 \times 3$ matrix, filled row-wise. Then:

```{r}
B <- matrix(A_1, nrow=2, ncol=3, byrow=TRUE)
```

#### (i) Replace any elements that are equal to 8, with the square of the element at $(1, 2)$ of (b)

```{r}
B[B == 8] <- B[1, 2] ^ 2
B
```

#### (ii) Confirm that all values of (b) are less than or equal to $25$ AND greater than $4$

```{r}
all((B <= 25) & (B > 4))
```

### (c) Create a $3 \times 2 \times 3$ array, using the following vector: `c( 10, 5, 1, 4, 7, 4, 3, 3, 1, 3, 4, 3, 1, 7, 8, 3, 7, 3)`. Then:

```{r}
C <- array(
        c(10, 5, 1, 4, 7, 4, 3, 3, 1, 3, 4, 3, 1, 7, 8, 5, 3, 7, 3),
        dim=c(3, 2, 3)
        )
```

#### (i) Identify the dimension-specific index positions of elements that are either $3$ OR $4$

```{r}
which((C == 3) | (C == 4), arr.ind=TRUE)
```

#### (ii) Replace all elements in (c) that are less than $3$ OR greater than or equal to $7$ with $100$

```{r}
C[(C < 3) | (C >= 7)] <- 100
C
```

### (d) Using (a), and the vector `c(F, T)` tot extract every second value. Can you do the same with the vector `c(1,0)`, why / why not?

```{r}
print(A[c(FALSE, TRUE)])
print(A[c(0, 1)])
# can't use 0/1 for FALSE/TRUE for indexing
```

# 4.2 Characters

## 4.2.1 Creating a String

Strings created between a set of quotes `""` or `''`

```{r}
foo <- "This is a character string!"
foo
```

R counts a string as a single entity

```{r}
length(x=foo)
```

Find string length using `nchar()`

```{r}
nchar(x=foo)
```

Can have any combination of chars in a string

```{r}
bar <- "23.3"
bar
```

Strings have no numeric meaning

```{r}
bar * 2
```

Can compare strings

```{r}
"alpha" == "alpha"
```

```{r}
"alpha" != "beta"
```

```{r}
c("alpha", "beta", "gamma") == "beta"
```

Letters later in the alphabet are greater than those near the start

```{r}
"alpha" <= "beta"
```

```{r}
"gamma" > "Alpha"
```

Uppercase letters are greater than lowercase

```{r}
"Alpha" > "alpha"
```

```{r}
"beta" >= "bEtA"
```

Most symbols are valid chars

```{r}
baz <- "&4 _ 3 **%.? $ymbolic non$en$e ,; "
baz
```

## 4.2.2 Concatenation

Use `cat()` to concatenate and print strings

```{r}
qux <- c("awsome", "R", "is")
length(x=qux)
```

```{r}
qux
```

```{r}
cat(qux[2], qux[3], "totally", qux[1], "!")
```

Use `paste()` to return a concatenated string

```{r}
paste(qux[2], qux[3], "totally", qux[1], "!")
```

Both have an optional keyword `sep` for specifying the separator

```{r}
paste(qux[2], qux[3], "totally", qux[1], "!", sep="---")
```

```{r}
paste(qux[2], qux[3], "totally", qux[1], "!", sep="")
```

```{r}
cat("Do you think ", qux[2], " ", qux[3], " ", qux[1], "?", sep="")
```

R will convert most types to strings

```{r}
a <- 3
b <- 4.4
cat("The value stored as 'a' is ", a, ".", sep="")
```

```{r}
paste("The value stored as 'b' is ", b, ".", sep="")
```

```{r}
cat("The result of a + b is ", a, " + ", b, " = ", a + b, sep="")
```

```{r}
paste(
    "Is ", a + b, " less than 10? That's totally ", 
    (a + b) < 10, ".", sep=""
    )
```

## 4.2.3 Escape Sequences

Standard escape characters:

-   \\n -\> newline

-   \\t -\> tab

-   \\b -\> backspace

-   \\\\ -\> backslash

-   \\" -\> double quote

```{r}
cat("here is a string\nsplit\tto neww\b \n\n\tlines")
```

```{r}
cat("I really want a backslash: \\\nand a double quote\"")
```

## 4.2.4 Substrings and Matching

Use `substr()` to get parts of a string

```{r}
foo <-"This is a character string!"
foo
```

```{r}
substr(x=foo, start=21, stop=27)
```

Can be used to reassign parts of strings

```{r}
substr(x=foo, start=1, stop=4) <- "Here"
foo
```

Can use `sub()` to replace a pattern or `gsub()` to replace all matches

```{r}
bar <- "How much wood could a woodchuck chuck"
sub(pattern="chuck", replacement="hurl", x=bar)
```

```{r}
gsub(pattern="chuck", replacement="hurl", x=bar)
```

## 4.4 Exercises

### (a) Recreate the following:

    "The quick brown fox
        jumped over
            the lazy dogs"

```{r}
cat('"The quick brown fox\n\tjumped over\n\t\tthe lazy dogs"')
```

### (b) Store `num1 <- 4` and `num2 <- 0.75`. Recreate:

    [1] "The result of multiplying 4 by 0.75 is 3"

```{r}
num1 <- 4
num2 <- 0.75

paste('The result of multiplying ', num1, ' by ', num2, ' is ', num1 * num2)
```

### (c) Replace \${user} in working directory with your initial and surname

```{r}
directory <- getwd()
sub(pattern='jagma', replacement='gallen', directory)
```

### (d) Store `"How much wood could a woodchuck chuck"`. Then:

```{r}
bar <- "How much wood could a woodchuck chuck"
bar
```

#### (i) Store a new string of (d) plus `"if a woodchuck could chuck wood"`

```{r}
baz <- paste(bar, "if a woodchuck could chuck wood")
baz
```

#### (ii) From (i) replace all instance of `wood` with `metal`

```{r}
gsub(pattern='wood', replacement='metal', baz)
```

### (e) Store `"Two 6-packs for $12.99"`. Then:

```{r}
E <- "Two 6-packs for £12.99"
E
```

#### (i) Use a check for equality to confirm the substring from $5$ to $10$ is `"6-pack"`

```{r}
substr(E, start=5, stop='10') == '6-pack'
```

#### (ii) Make it a better deal, by changing the price to $\$10.99$

```{r}
E <- sub(pattern='12', replacement='10', E)
E
```

# 4.3 Factors

Useful for categorical data

## 4.3.1 Identifying Categories

Data:

| Person   | Sex    | Month of Birth |
|----------|--------|----------------|
| Liz      | Female | April          |
| Jolene   | Female | January        |
| Susan    | Female | December       |
| Boris    | Male   | September      |
| Rochelle | Female | November       |
| Tim      | Male   | July           |
| Simon    | Male   | July           |
| Amy      | Female | June           |

Save names as a vector

```{r}
first_name <- c(
                "Liz", "Jolene", "Susan", "Boris", "Rochelle",
                "Tim", "Simon", "Amy"
                )
```

Sex is categorical, limited number of options

```{r}
sex.num <- c(0, 0, 0, 1, 0, 1, 1, 0)
sex.char <- c(
                "female", "female", "female", "male", "female", 
                "male", "male", "female"
                )
```

Use `factor()` to convert vector to a factor

```{r}
sex.num.fac <- factor(x=sex.num)
sex.num.fac
```

```{r}
sex.char.fac <- factor(x=sex.char)
sex.char.fac
```

Use `levels()` to obtain factor level information

```{r}
levels(x=sex.num.fac)
```

```{r}
levels(x=sex.char.fac)
```

Can relabel factor levels

```{r}
levels(x=sex.num.fac) <- c("1", "2")
sex.num.fac
```

Subset like other vectors

```{r}
sex.char.fac[2:5]
```

```{r}
sex.char.fac[c(1:3, 5, 8)]
```

Need to use a string to use logical comparators

```{r}
sex.num.fac == "2"
```

Can use corresponding vector positions for subsetting

```{r}
first_name[sex.char.fac == "male"]
```

## 4.3.2 Defining and Ordering Levels

Use month of birth

```{r}
mob <- c("Apr", "Jan", "Dec", "Sep", "Nov", "Jul", "Jul", "Jun")
```

Unordered currently

```{r}
mob[2]
```

```{r}
mob[3]
```

```{r}
mob[2] < mob[3]
```

Create an ordered factor with `ordered` keyword, with explicit `levels`

```{r}
ms <- c(
        "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
        "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
        )
mob.fac <- factor(x=mob, levels=ms, ordered=TRUE)
mob.fac
```

Now have meaningful comparisons

```{r}
mob.fac[2] < mob.fac[3]
```

## 4.3.3 Combining and Cutting

Simple vectors

```{r}
foo <- c(5.1, 3.3, 3.1, 4)
bar <- c(4.5, 1.2)
c(foo, bar)
```

`c()` on factors

```{r}
new.values <- factor(
                    x=c("Oct", "Feb", "Feb"),
                    levels=levels(mob.fac),
                    ordered=TRUE
                     )
new.values
```

*Used to return just a numeric vector*

```{r}
c(mob.fac, new.values)
```

*Old method for combining*

```{r}
mob.new <- levels(mob.fac)[c(mob.fac, new.values)]
mob.new.fac <- factor(x=mob.new, levels=levels(mob.fac), ordered=TRUE)
mob.new.fac
```

Can `cut()` vectors to produce intervals

```{r}
Y <- c(0.53, 5.4, 1.5, 3.33, 0.45, 0.01, 2, 4.2, 1.99, 1.01)
br <- c(0, 2, 4, 6)
cut(x=Y, breaks=br)
```

Can reverse the interval boundaries using the `right` keyword

```{r}
cut(x=Y, breaks=br, right=F)
```

Can include lowest or highest value with `include.lowest` depending on `right`

```{r}
cut(x=Y, breaks=br, right=F, include.lowest=T)
```

Can also change the `levels` of this with `labels`

```{r}
lab <- c("Small", "Medium", "Large")
cut(x=Y, breaks=br, right=F, includ.lowest=T, labels=lab)
```

## 4.5 Exercises

### New Zealand Government

-   Parties

    -   National

    -   Labour

    -   Greens

    -   Māori

    -   Other

Ask 20 New Zealanders about political persuasion

-   12 Males, 8 Females

    -   Numbers 1, 5-7, 12, and 14-16 being female

-   Individuals

    -   1, 4, 12, 15, 16, and 19 Labour

    -   No Māori

    -   6, 9, and 11 Greens

    -   10, and 20 Other

    -   Rest National

### (a) Create Sex and Party vectors

```{r}
sex <- 1:20
sex[c(1, 5:7, 12, 14:16)] <- 'F'
sex[c(2:4, 8:11, 13, 17:20)] <- 'M'
sex
```

```{r}
party <- 1:20
party[c(1, 4, 12, 15, 16, 19)] <- 'Labour'
party[c(6, 9, 11)] <- 'Greens'
party[c(10, 20)] <- 'Other'
party[c(2, 3, 5, 7, 8, 13, 14, 17, 18)] <- 'National'
party
```

### (b) Create factors from (a)

```{r}
sexes <- c('M', 'F')
sex.fac <- factor(sex, levels=sexes)
sex.fac
```

```{r}
parties <- c('Labour', 'Greens', 'National', 'Maori', 'Other')
party.fac <- factor(party, levels=parties)
party.fac
```

### (c) Use (b) and subsetting to:

#### (i) Return parties for only males

```{r}
party.fac[sex.fac == 'M']
```

#### (ii) Return genders of National party

```{r}
sex.fac[party.fac == 'National']
```

### (d) Another six people:

-   `c("National", "Maori", "Maori", "Labour", "Greens", "Labour")`

-   `c("M", "M", "F", "F", "F", "M")`

```{r}
party.new <- c(
                'National', 'Maori', 'Maori', 
                'Labour', 'Greens', 'Labour'
               )
party.new.fac <- factor(c(party, party.new), 
                        levels=levels(party.fac))
party.new.fac
```

```{r}
sex.new <- c('M', 'M', 'F', 'F', 'F', 'M')

sex.new.fac <- factor(c(sex, sex.new), 
                        levels=levels(sex.fac))
sex.new.fac
```

### Ask individuals about election confidence percentage:

Results:

$93, 55, 29, 100, 52, 84, 56, 0, 33, 52, 35, 53, 55,$

$46, 40, 40, 56, 45, 64, 31, 10, 29, 40, 95, 18, 61$

### (e) Create factor with following levels:

-   Low: $[0, 30]$

-   Moderate: $(30, 70]$

-   High: $(70, 100]$

```{r}
percentages <- c(
                93, 55, 29, 100, 52, 84, 56, 0, 33, 52, 35, 53, 55,
                46, 40, 40, 56, 45, 64, 31, 10, 29, 40, 95, 18, 61
                )

cuts <- c(0, 30, 70, 100)
labs <- c('Low', 'Moderate', 'High')

confidence <- cut(percentages, breaks=cuts, include.lowest=TRUE, labels=labs)
confidence
```

### (f) From (e) extract levels corresponding to Labour and National

```{r}
confidence.labour <- confidence[party.new.fac == 'Labour']
confidence.labour
```

```{r}
confidence.national <- confidence[party.new.fac == 'National']
confidence.national
```
