---
title: "Chapter 2 Numerics, Arithmetic, Assignment, and Vectors"
output: html_notebook
---

# 2.1 Basic Maths

## 2.1.1 Arithmetic

BIDMAS is important, use brackets to distinguish

```{r}
2 + 3
14 / 6
14 / 6 + 5
14 / (6 + 5)
3 ^ 2
2 ^ 3
```

Can find square root of any non-negative number

```{r}
sqrt(x=9)
sqrt(x=5.311)
```

## 2.1.2 Logs and Exponents

Both x and the base must be positive

```{r}
log(x=243, base=3)
```

Base e exponent and log base e by default

```{r}
exp(x=3)
```

```{r}
log(x=20.08554)
```

## 2.1.3 E-Notation

Large and very small numbers will use E-notation instead of x10\^x

```{r}
2342151012900
0.0000002533
```

## 2.1 Exercises

### (a) Verify $\frac {6a + 42} {3^{4.2-3.62}} = 29.50556$ when $a=2.3$

```{r}
a <- 2.3
((6 * a) + 42) / (3 ^ (4.2 - 3.62))
```

### (b) Which of the following squares $-4$ and adds $2$?

#### (i) (-4) \^ 2 + 2 (ii) -4 \^ 2 + 2 (iii) (-4) \^( 2 + 2) (iv) -4 \^( 2 + 2)

```{r}
(-4) ^ 2 + 2
```

### (c) What is the square root of the average of the numbers: $25.2, 15, 16.44, 15.3,18.6$ ?

```{r}
sqrt((mean(c(25.2, 15, 16.44, 15.3, 18.6))) / 2)
```

### (d) Find $\log_e 0.3$

```{r}
log(0.3)
```

### (e) Compute the exponential of (d)

```{r}
exp(-1.203973)
```

### (f) What is R's representation of $-0.00000000423546322$?

```{r}
-0.00000000423546322
```

# 2.2 Assigning Objects

Can assign objects using `=` or `<-`

```{r}
x <- -5
x
x = x + 1
x
```

```{r}
my_number = 45.2
y <- my_number * x
y
```

```{r}
ls()
```

## 2.2 Exercises

### (a) Create an object that stores $3^{2} \times 4^{\frac {1} {8}}$

```{r}
a <- (3 ^ 2) * (4 ^ (1 / 8))
print(a)
```

### (b) Overwrite (a) by dividing it by $2.33$

```{r}
a <- a / 2.33
print(a)
```

### (c) Create a new object with the value $-8.2 \times 10^{-13}$

```{r}
c <- -8.2e-13
print(c)
```

### (d) Print (c) times (b)

```{r}
d <- a * c
print(d)
```

# 2.3 Vectors

## 2.3.1 Creating Vectors

Create vectors using the `c()` function

```{r}
my_vec <- c(1, 3, 1, 42)
my_vec
```

Can have calculated values within the construction of the vector

```{r}
foo <- 32.1
my_vec_2 <- c(3, -3, 2, 3.45, 1e+03, 64 ^ 0.5, 2 + (3 - 1.1) / 9.44, foo)
my_vec_2
```

Can make vectors from other objects

```{r}
my_vec_3 = c(my_vec, my_vec_2)
my_vec_3
```

## 2.3.2 Sequences, Repetition, Sorting, and Lengths

Create sequences with the `:` operator

```{r}
3:27
```

Can use calculated values as well

```{r}
foo <- 5.3
bar <- foo:(-47 + 1.5)
bar
```

### Sequences with `seq`

Can also create sequences with `seq()`

```{r}
seq(from=3, to=27, by=3)
```

Can define the separation with `by` keyword, or the length of the sequence with `length.out`

```{r}
seq(from=3, to=27, length.out=40)
```

Can use calculated fields and negatives for decreasing sequences

```{r}
foo <- 5.3
my_seq <- seq(from=foo, to=(-47 + 1.5), by=-2.4)
my_seq
```

### Repetition with `rep`

Repeat values multiple `times` over

```{r}
rep(x=1, times=4)
```

```{r}
rep(x=c(3, 62, 8.3), times=3)
```

Repeat `each` element of a vector

```{r}
rep(x=c(3, 62, 8.3), each=2)
```

`each` occurs first then `times`

```{r}
rep(x=c(3, 62, 8.3), times=3, each=2)
```

Can combine all of the above into a single vector

```{r}
foo <- 4 
c(3, 8.3, rep(x=32, times=foo), seq(from=-2, to=1, length.out=(foo + 1)))
```

### Sorting with `sort`

Sort values into ascending order

```{r}
sort(x=c(2.5, -1, -10.3, 3.44), decreasing=FALSE)
```

Or descending order

```{r}
sort(x=c(2.5, -1, -10.3, 3.44), decreasing=TRUE)
```

```{r}
foo <- seq(from=4.3, to=5.5, length.out=8)
foo
```

```{r}
bar <- sort(x=foo, decreasing=TRUE)
bar
```

```{r}
sort(x=c(foo, bar), decreasing=FALSE)
```

### Finding Vector Length with `length`

Finds the number of entries in a vector

```{r}
length(x=c(3, 2, 8, 1))
```

```{r}
length(x=5:13)
```

```{r}
foo <- 4
bar <- c(3, 8.3, 
         rep(x=32, times=foo), 
         seq(from=-2, to=1, length.out=foo+1))
length(x=bar)
```

## 2.3 Exercises

### (a) Store values from $5$ to $-11$ in steps of $0.3$

```{r}
a <- seq(from=5, to=-11, by=-0.3)
print(a)
```

### (b) Overwrite (a) with the same sequence in revered order

```{r}
a <- sort(a, decreasing=FALSE)
print(a)
```

### (c) Repeat the vector `c(-1, 3, -5, 7, -9)` twice, with each element repeated 10 times. Display the result in descending order.

```{r}
c <- rep(c(-1.3, -5.7, -9), times=2, each=10)
print(sort(c, decreasing=TRUE))
```

### (d) Create and store a vector that contains the following:

#### (i) Sequence of integers from $6$ to $12$

```{r}
d_i <- 6:12
d_i
```

#### (ii) Threefold repetition of $5.3$

```{r}
d_ii <- rep(5.3, times=3)
d_ii
```

#### (iii) The number $-3$

```{r}
d_iii <- -3
d_iii
```

#### (iv) A sequence of nine values from $102$ and ending at the length of (c)

```{r}
d_iv <- seq(from=102, to=length(c), length.out=9)
d_iv
```

```{r}
d <- c(d_i, d_ii, d_iii, d_iv)
print(d)
```

### (e) Confirm the length of (d) is $20$

```{r}
print(length(d))
```

## 2.3.2 Sub-setting and Element Extraction

Can index vectors with `[]`

```{r}
my_vec <- c(5, -2.3, 4, 4, 4, 6, 8, 10, 40221, -8)
length(x=my_vec)
```

R is 1 indexed

```{r}
my_vec[1]
```

```{r}
foo <- my_vec[2]
foo
```

Going from 1 - N inclusive

```{r}
my_vec[length(x=my_vec)]
```

```{r}
my_vec.len <- length(x=my_vec)
bar <- my_vec[my_vec.len - 1]
bar
```

All possible indices

```{r}
1:my_vec.len
```

Delete elements using negative indices, doesn't overwrite original object

```{r}
my_vec[-1]
```

```{r}
baz <- my_vec[-2]
baz
```

```{r}
qux <- my_vec[-(my_vec.len - 1)]
qux
```

Can piece it back together from `bar` and `qux`

```{r}
c(qux[-length(x=qux)], bar, qux[length(x=qux)])
```

Can subset vectors with other vectors

```{r}
my_vec[c(1, 3, 5)]
```

```{r}
foo <- my_vec[1:4]
foo
```

Order of indices matters for the returned values

```{r}
foo[length(foo):2]
```

Can use any amount of indices as long as they are within the range `1:length(v)`

```{r}
indexes <- c(4, rep(x=2, times=3), 1, 1, 2, 3:1)
indexes
```

```{r}
foo[indexes]
```

Can also be used to remove multiple values

```{r}
foo[-c(1, 3)]
```

Not possible to mix positive and negative indices

Can overwrite values with indexing

```{r}
bar <- c(3, 2, 4, 4, 1, 2, 4, 1, 0, 0, 5)
bar
```

```{r}
bar[1] <- 6
bar
```

```{r}
bar[c(2, 4, 6)] <- c(-2, -0.5, -1)
bar
```

```{r}
bar[7:10] <- 100
bar
```

## 2.4 Exercises

### (a) Create a vector that stores:

-   A sequence of length $5$ from $3$ to $6$

-   A twofold repetition of the vector `c(2, -5.1, -33)`

-   The value $\frac {7} {42} + 2$

```{r}
a <- c(seq(from=3, to=6, length.out=5), 
       rep(c(2, -5.1, -33), times=2), (7 / 12) + 2)
print(a)
```

### (b) Extract the first and last elements from (a)

```{r}
b <- c(a[1], a[length(a)])
print(b)
```

### (c) Store everything from (a) except the first and last element

```{r}
c <- a[c(-1, -length(a))]
print(c)
```

### (d) Use (b) and (c) to reconstruct (a)

```{r}
d <- c(b[1], c, b[length(b)])
print(d)
```

### (e) Overwrite (a) with the same values in ascending order

```{r}
a <- sort(a, decreasing=FALSE)
print(a)
```

### (f) Use `:` to reverse the order of (e)

```{r}
f <- a[length(a):1]  # reverses the sorted vector
print(f)
```

### (g) Create a vector from (c) that repeats the third element of (c) three times, the sixth four times, and the last once

```{r}
g <- c(rep(c[3], times=3), rep(c[6], times=4), c[length(c)])
print(g)
```

### (h) Create a copy of (e), overwrite the first, fifth to the seventh, and last elements with the values 99 to 95.

```{r}
h <- a
h[c(1, 5:7, length(h))] <- 99:95
print(h)
```

## 2.3.4 Vector-Oriented Behaviour

Element-wise operations

```{r}
foo <- 5.5:0.5
foo
```

```{r}
foo - c(2, 4, 6, 8, 10, 12)
```

Smaller vectors are repeated to the size of the larger

```{r}
bar <- c(1, -1)
foo * bar
```

Non-multiples will produce a warning, but will partially repeat the smaller

```{r}
baz <- c(1, -1, 0.5, -0.5)
foo * baz
```

Scalars will also do element wise operations

```{r}
qux <- 3
foo + qux
```

Can use aggregate functions on vectors

```{r}
sum(foo)
```

```{r}
prod(foo)
```

The repetitions also works with assignment

```{r}
foo[c(1, 3, 5, 6)] <- c(-99, 99)
foo
```

## 2.5 Exercises

### (a) Convert the vector `c( 2,0.5,1,2,0.5,1,2,0.5,1)` to only 1s, using a vector of length 3

```{r}
a <- c(2, 0.5, 1, 2, 0.5, 1, 2, 0.5, 1)
print(a)
a <- a / c(2, 0.5, 1)
print(a)
```

### (b) $^{\circ}C = \frac {5}{9} (^{\circ}F - 32)$ Use vectors to convert $(45, 77, 20, 19, 101, 120, 212)$ from Fahrenheit to Celsius

```{r}
F <- c(45, 77, 20, 19, 101, 120, 212)
C <- (5 / 9) * (F - 32)
print(C)
```

### (c) Use the vector `c(2, 4, 6)` and `c(1, 2)` with `rep` and `*` to produce `c(2, 4, 6, 4, 8, 12)`

```{r}
c_1 <- c(2, 4, 6)
c_2 <- c(1,2)
c_3 <- rep(c_1, times=2) * rep(c_2, each=3)
print(c_3)
```

### (d) Overwrite the middle four elements of (c) with the two values of $-0.1$ and $-100$

```{r}
c_3[2:(length(c_3) - 1)] <- c(-0.1, 100)
print(c_3)
```
