---
title: "Chapter 6 Special Values, Classes, and Coercion"
output: html_notebook
---

# 6.1 Some Special Values

## 6.1.1 Infinity

Reserved keyword `Inf`

```{r}
foo <- Inf
foo
```

```{r}
bar <- c(3401, Inf, 3.1, -555, Inf, 43)
bar
```

Numbers too large will result in `Inf`

```{r}
baz <- 90000 ^ 100
baz
```

Can also represent `-Inf`

```{r}
qux <- c(-42, 565, -Inf, -Inf, Inf, -45632.3)
qux
```

`Inf` times any negative value is `-Inf`, any other arithmetic results in $\pm$`Inf`

```{r}
Inf * -9
```

```{r}
Inf + 1
```

```{r}
4 * -Inf
```

```{r}
-45.2 - Inf
```

```{r}
Inf - 45.2
```

```{r}
Inf + Inf
```

```{r}
Inf / 23
```

Dividing by `Inf` will result in $0$

```{r}
-59 / Inf
```

```{r}
-59 / -Inf
```

Any non-zero value divided by $0$ will result in `Inf`

```{r}
-59 / 0
```

```{r}
59 / 0
```

```{r}
Inf / 0
```

Can check for `Inf` with either `is.infinite()` or `is.finite()`

```{r}
qux
```

```{r}
is.infinite(x=qux)
```

```{r}
is.finite(x=qux)
```

Relational operators function as expected

```{r}
-Inf < Inf
```

```{r}
Inf > Inf
```

```{r}
qux == Inf
```

```{r}
qux == -Inf
```

## 6.1.2 NaN

Unquantifiable values are give `NaN` or Not a Number values

```{r}
foo <- NaN
foo
```

```{r}
bar <- c(NaN, 54.3, -2, NaN, 90094.123, -Inf, 55)
bar
```

Adding `Inf` and `-Inf` of diving `Inf` by `Inf` will also result in `NaN`

```{r}
-Inf + Inf
```

```{r}
Inf / Inf
```

$\frac {0}{0}$ will also return `NaN`

```{r}
0 / 0
```

Any arithmetic with `NaN` will also result in `NaN`

```{r}
NaN + 1
```

```{r}
2 + 6 * (4 - 4) / 0
```

```{r}
3.5 ^ (-Inf / Inf)
```

Can test for `NaN` with `is.nan()`

```{r}
bar
```

```{r}
is.nan(x=bar)
```

Can always use negation to get non-NaN values

```{r}
!is.nan(x=bar)
```

```{r}
is.nan(x=bar) | is.infinite(x=bar)
```

```{r}
bar[-(which(is.nan(x=bar) | is.infinite(x=bar)))]
```

## 6.1 Exercises

### (a) Store the following:

    foo <- c(13563, -14156, -14319, 16981, 12921, 11979, 9568, 8833, -12968, 8133)

```{r}
foo <- c(13563, -14156, -14319, 16981, 12921, 11979, 9568, 8833, -12968, 8133)
```

#### (i) Output all elements when raised to the power of $75$ are NOT infinite

```{r}
foo[is.finite(foo ^ 75)]
```

#### (ii) Return elements of `foo`, that when raised to the power of $75$ are NOT negative infinity

```{r}
foo[foo ^ 75 != -Inf]
```

### (b) Store the following $3 \times 4$ matrix

$$
\left[ \begin{array}{cc}
77875.40 & 27551.45 & 23764.30 & -36478.88 \\
-35466.25 & -73333.85 & 36599.69 & -70585.69 \\
-39803.81 & 55976.34 & 76694.82 & 47032.00
\end{array}\right]
$$

```{r}
bar <- matrix(
            c(77875.4, -35466.25, -39803.81, 
              27551.45, -73333.85, 55976.34,
              23764.3, 36599.69, 76694.82, 
              -36478.88, -70585.69, 47032),
            nrow=3, ncol=4
            )
```

#### (i) Identify the coordinate-specific indices of the entries that are `NaN` when you raise them to the power of $65$ and divide by infinity

```{r}
which(is.nan((bar ^ 65) / Inf), arr.ind=TRUE)
```

#### (ii) Return the values that are NOT `NaN` when you raise them to the power of $67$ and infinity is added. Confirm that this is equal to those that are raised to the power of $67$ are not negative infinity

```{r}
B_1 <- bar[!is.nan((bar ^ 67) + Inf)]
B_1
```

```{r}
B_2 <- bar[bar ^ 67 != -Inf]
B_2
```

```{r}
all(B_1 == B_2)
```

#### (iii) Identify values that are either negative infinity OR finite when raised to the $67^{th}$ power

```{r}
bar[(is.finite(bar ^ 67)) | (bar ^ 67 == -Inf)]
```

## 6.1.3 NA

Non-numerical version of `NaN` -\> `NA` for Not Available

```{r}
foo <- c("character", "a", NA, "with", "string", NA)
foo
```

```{r}
bar <- factor(
            c(
                "blue", NA, NA, "blue", 
                "green", "blue", NA, "red", 
                "red", NA, "green"
            )
            )
bar
```

```{r}
baz <- matrix(c(1:3, NA, 5, 6, NA, 8, NA), nrow=3, ncol=3)
baz
```

Can identify `NA` with `is.na()` , also identifies `NaN`

```{r}
qux <- c(
        NA, 5.89, Inf, NA, 9.43,
        -2.35, NaN, 2.10, -8.53, -7.58, 
        NA, -4.58, 2.01, NaN
        )
qux
```

```{r}
is.na(x=qux)
```

Need to combine with `is.nan()` to identify just `NA`

```{r}
which(x=is.nan(x=qux))
```

```{r}
which(x=(is.na(x=qux) & !is.nan(x=qux)))
```

Can easily remove `NA` with `na.omit()`

```{r}
quux <- na.omit(object=qux)
quux
```

Similar to `NaN` any operations with `NA` produces `NA`

```{r}
3 + 2.1 * NA - 4
```

```{r}
3 * c(1, 2, NA, NA, NaN, 6)
```

```{r}
NA > 76
```

```{r}
76 > NaN
```

## 6.1.4 NULL

Used for empty values instead of missing values

```{r}
foo <- NULL
foo
```

```{r}
bar <- NA
bar
```

`NULL` is empty and doesn't print with a vector element, also applies to making vectors

```{r}
c(2, 4, NA, 8)
```

```{r}
c(2, 4, NULL, 8)
```

`NULL` cannot take up a vector position

```{r}
c(NA, NA, NA)
```

```{r}
c(NULL, NULL, NULL)
```

Can check if an object is `NULL` with `is.null()` , returns only a single bool

```{r}
opt.arg <- c("string1", "string2", "string3")
opt.arg
```

```{r}
is.na(x=opt.arg)
```

```{r}
is.null(x=opt.arg)
```

```{r}
opt.arg <- c(NA, NA, NA)
is.na(x=opt.arg)
```

```{r}
opt.arg <- c(NULL, NULL, NULL)
is.null(x=opt.arg)
```

Inclusion of `NULL` in arithmetic results in an empty object

```{r}
NULL + 53
```

```{r}
53 <= NULL
```

```{r}
NaN - NULL + NA / Inf
```

`NULL` will be returned when invalid calls are made

```{r}
foo <- list(member1=c(33, 1, 5.2, 7), member2="NA or NULL?")
foo
```

```{r}
foo$member3
```

Can fill empty members with anything

```{r}
foo$member3 <- NA
foo
```

## 6.2 Exercises

### (a) Consider the following:

    foo <- c(4.3, 2.2, NULL, 2.4, NaN, 3.3, 3.1, NULL, 3.4, NA)

```{r}
foo <- c(4.3, 2.2, NULL, 2.4, NaN, 3.3, 3.1, NULL, 3.4, NA)
```

Which of the following are true?

#### (i) The length of `foo` is $8$ 

```{r}
length(foo)
```

#### (ii) Calling `which(x=is.na(x=foo))` will not result in $4$ and $8$ 

```{r}
which(is.na(foo))
```

#### (iii) Checking `is.null(x=foo)` will provide the locations of `NULL`

```{r}
is.null(foo)
```

#### (iv) Executing `is.na(x=foo[8]) + 4 / NULL` will not result in `NA`

```{r}
is.na(foo[8]) + 4 / NULL
```

### (b) Store a list containing:

    c(7, 7, NA, 3, NA, 1, 1, 5, NA)

```{r}
B <- list(c(7.7, NA, 3, NA, 1, 1, 5, NA))
```

#### (i) Name the member `"alpha"`

```{R}
names(B) <- 'alpha'
```

#### (ii) Confirm the list doesn't have a member named `"beta"`

```{r}
is.null(B$beta)
```

#### (iii) Add a new member called `beta`, which is the vector obtained from identifying the indices of `NA` in `alpha`

```{r}
B$beta <- which(is.na(B$alpha))
B
```

# 6.2 Understanding Types, Classes, and Coercion

## 6.2.1 Attributes

Can reveal a classes attributes using `attributes()`

```{r}
foo <- matrix(data=1:9, nrow=3, ncol=3)
foo
```

```{r}
attributes(foo)
```

Can retrieve contents of an attribute with `attr()`

```{r}
attr(x=foo, which="dim")
```

```{r}
bar <- matrix(
            data=1:9, 
            nrow=3, 
            ncol=3, 
            dimnames=list(
                        c("A", "B", "C"),
                        c("D", "E", "F")
                        )
            )
bar
```

```{r}
attributes(bar)
```

Can extract values from attributes with member extraction, or specific functions for common attributes

```{r}
dimnames(bar)
```

Can overwrite some attributes

```{r}
dimnames(foo) <- list(c("A", "B", "C"), c("D", "E", "F"))
foo
```

## 6.2.2 Object Class

### Stand-Alone Vectors

```{r}
num.vec1 <- 1:4
num.vec1
```

```{r}
num.vec2 <- seq(from=1, to=4, length=6)
num.vec2
```

```{r}
char.vec <- c("a", "few", "strings", "here")
char.vec
```

```{r}
logic.vec <- c(T, F, F, F, T, F, T, T)
logic.vec
```

```{r}
fac.vec <- factor(
            c("Blue", "Blue", "Green", "Red", "Green", "Yellow")
                )
fac.vec
```

Can determine the class of an object with the `class()` function

```{r}
class(num.vec1)
```

```{r}
class(num.vec2)
```

```{r}
class(char.vec)
```

```{r}
class(logic.vec)
```

```{r}
class(fac.vec)
```

### Other Data Structures

`class()` returns the nature of the object, not the type

```{r}
num.mat1 <- matrix(data=num.vec1, nrow=2, ncol=2)
num.mat1
```

```{r}
num.mat2 <- matrix(data=num.vec2, nrow=2, ncol=3)
num.mat2
```

```{r}
char.mat <- matrix(data=char.vec, nrow=2, ncol=2)
char.mat
```

```{r}
logic.mat <- matrix(data=logic.vec, nrow=4, ncol=2)
logic.mat
```

```{r}
class(num.mat1)
```

```{r}
class(num.mat2)
```

```{r}
class(char.mat)
```

```{r}
class(logic.mat)
```

### Multiple Classes

Objects can be part of multiple classes

```{r}
ordfac.vec <- factor(
                x=c("Small", "Large", "Large", "Regular", "Small"),
                levels=c("Small", "Regular", "Large"),
                ordered=TRUE
                    )
ordfac.vec
```

```{r}
class(ordfac.vec)
```

## 6.2.3 Is-Dot Object-Checking Functions

Can use `is.<class | type>()` functions to check objects

```{r}
num.vec1 <- 1:4
num.vec1
```

Type versions

```{r}
is.integer(num.vec1)
```

```{r}
is.numeric(num.vec1)
```

```{r}
is.logical(num.vec1)
```

Class versions

```{r}
is.matrix(num.vec1)
```

```{r}
is.data.frame(num.vec1)
```

```{r}
is.vector(num.vec1)
```

## 6.2.4 As-Dot Coercion Functions

Many objects can be coerced into other types using `as.<class | type>()`

Some types are automatically coerced depending on the operation

```{r}
1:4 + c(T, F, F, T)
```

```{r}
foo <- 34
bar <- T
paste(
    "Definitely foo: ", foo, "; definitely bar: ", bar, ".", sep=''
    )
```

Can us `as.<class | type>()` to explicitly cast objects

```{r}
as.numeric(c(T, F, F, T))
```

```{r}
1:4 + as.numeric(c(T, F, F, T))
```

```{r}
foo.ch <- as.character(foo)
foo.ch
```

```{r}
bar.ch <- as.character(bar)
bar.ch
```

```{r}
paste(
    "Definitely foo: ", foo.ch, 
    "; definitely bar: ", bar.ch, ".", sep=''
    )
```

Coercion and casting can only happen when they make sense

```{r}
as.numeric("32.4")
```

```{r}
as.numeric("g'day mate")
```

Direct Char -\> Logical is not possible

```{r}
as.logical(c("1", "0", "1", "0", "0"))
```

```{r}
as.logical(as.numeric(c("1", "0", "1", "0", "0")))
```

Some casts are not intuitive

```{r}
baz <- factor(x=c("male", "male", "female", "male"))
baz
```

```{r}
as.numeric(baz)
```

Factors are numeric under the levels

```{r}
qux <- factor(x=c(2, 2, 3, 5))
qux
```

```{r}
as.numeric(qux)
```

Object casting

```{r}
foo <- matrix(data=1:4, nrow=2, ncol=2)
foo
```

```{r}
as.vector(foo)
```

Column-wise deconstruction

```{r}
bar <- array(
            data=c(8, 1, 9, 5, 5, 1, 3, 4, 3, 9, 8, 8), 
            dim=c(2, 3, 2)
            )
bar
```

```{r}
as.matrix(bar)
```

```{r}
as.vector(bar)
```

Objects of the wrong dimensions cannot be coerced

```{r}
baz <- list(
            var1=foo, 
            var2=c(T, F, T), 
            var3=factor(x=c(2, 3, 4, 4, 2))
            )
baz
```

```{r}
as.data.frame(baz)
```

```{r}
qux <- list(
            var1=c(3, 4, 5, 1),
            var2=c(T, F, T, T),
            var3=factor(x=c(4, 4, 2, 1))
            )
qux
```

```{r}
as.data.frame(qux)
```

## 6.3 Exercises

### (a) Identify the class of the following:

#### (i) `foo <- array(data=1:36, dim=c(3, 3, 4))`

```{r}
foo <- array(1:36, dim=c(3, 3, 4))
class(foo)
```

#### (ii) `bar <- as.vector(foo)`

```{r}
bar <- as.vector(foo)
class(bar)
```

#### (iii) `baz <- as.character(baz)`

```{r}
baz <- as.character(bar)
class(baz)
```

#### (iv) `qux <- as.factor(baz)`

```{r}
qux <- as.factor(baz)
class(qux)
```

#### (v) `quux <- bar + c(-0.1, 0.1)`

```{r}
quux <- bar + c(-0.1, 0.1)
class(quux)
```

### (b) For each object in (a), find the sum of the result of calling `is.numeric()` and `is.integer` separately 

```{r}
ai <- is.numeric(foo) + is.integer(foo)
aii <- is.numeric(bar) + is.integer(bar)
aiii <- is.numeric(baz) + is.integer(baz)
aiv <- is.numeric(qux) + is.integer(qux)
av <- is.numeric(quux) + is.integer(quux)

cat(
    "ai: ", ai, ", aii: ", aii, 
    ", aiii: ", aiii, ", aiv: ", aiv,
    ", av: ", av, sep=''
    )
```

#### Then turn the collection into a factor with levels $0$, $1$, and $2$ compare the factor with the coerced numeric vector

```{r}
B <- factor(c(ai, aii, aiii, aiv, av))
print(B)
print(as.numeric(B))
```

### (c) Turn the following:

            [,1]    [,2]    [,3]    [,4]
    [1,]    2       5       8       11  
    [2,]    3       6       9       12 
    [3,]    4       7       10      13

Into:

    [1] "2" "5" "8" "11" "3" "6" "9" "12" "4" "7" "10" "13"

```{r}
C <- matrix(2:13, nrow=3, ncol=4)
as.character(c(C[1,], C[2,], C[3,]))
```

### (d) Store the following matrix:

$$
\left[ \begin{array}{cc}
34 & 0 & 1 \\
23 & 1 & 2 \\
33 & 1 & 1 \\
42 & 0 & 1 \\
41 & 0 & 2
\end{array} \right]
$$

```{r}
D <- matrix(
            c(34, 0, 1,
              23, 1, 2,
              33, 1, 1,
              42, 0, 1,
              41, 0, 2),
            nrow=5, ncol=3, byrow=TRUE
            )
D
```

#### (i) Coerce the matrix to a data frame

```{r}
D.data.frame <- as.data.frame(D)
D.data.frame
```

#### (ii) As a data frame, coerce the second column to be logical

```{r}
D.data.frame[, 2] <- as.logical(D.data.frame[, 2])
D.data.frame
```

#### (iii) As a data frame, coerce the third column to be a factor

```{r}
D.data.frame[, 3] <- as.factor(D.data.frame[, 3])
D.data.frame
```
