---
title: "Chapter 5 Lists and Data Frames"
output: html_notebook
---

# 5.1 Lists of Objects

## 5.1.1 Definition and Component Access

`list()` can store any objects

```{r}
foo <- list(
            matrix(
                    data=1:4,
                    nrow=2,
                    ncol=2
                    ),
            c(T, F, T, T),
            "hello"
            )
foo
```

```{r}
length(x=foo)
```

Access element with `[[]]`

```{r}
foo[[1]]
```

```{r}
foo[[3]]
```

Can then access list elements like other objects

```{r}
foo[[1]] + 5.5
```

```{r}
foo[[1]][1, 2]
```

```{r}
foo[[1]][2,]
```

```{r}
cat(foo[[3]], "you!")
```

Can reassign elements like other objects

```{r}
foo[[3]]
```

```{r}
foo[[3]] <- paste(foo[[3]], "you!")
foo
```

Can access multiple elements as expected

```{r}
foo[[c(2, 3)]] # returns third element of second member
```

Instead need to use `[]` slicing

```{r}
bar <- foo[c(2, 3)]
bar
```

## 5.1.2 Naming

Can name elements of a list with `name()` function

```{r}
names(foo) <- c("my_matrix", "my_logicals", "my_string")
foo
```

Can now access list members using `$(member_name)`

```{r}
foo$my_matrix
```

Subsetting works just like before

```{r}
all(foo$my_matrix[, 2] == foo[[1]][, 2])
```

Can name members when list is being created

```{r}
baz <- list(
            tom=c(foo[[2]], T, T, T, F),
            dick="g'day mate",
            harry=foo$my_matrix * 2
            )
baz
```

## 5.1.3 Nesting

Can nest lists within lists, and assign new named members

```{r}
baz$bobby <- foo
baz
```

Can then access nested lists with another `$` or `[[]]`

```{r}
baz$bobby$my_logicals[1:3]
```

```{r}
baz[[4]][[2]][1:3]
```

Or even a mix

```{r}
baz[[4]]$my_logicals[1:3]
```

## 5.1 Exercises

### (a) Create a list that contains:

-   A sequence of $20$ evenly spaced numbers from $-4$ to $4$

-   A $3 \times 3$ matrix of the logical vector `c(F, T, T, T, F, T, T, F, F)` , column-wise filled

-   A char vector with the strings `"don"` and `"quixote"`

-   A factor containing `c("LOW", "MED", "LOW", "MED", "MED", "HIGH")`

```{r}
A <- list(
        seq(-4, 4, length.out=20),
        matrix(c(F, T, T, T, F, T, T, F, F), nrow=3, ncol=3),
        c('don', 'quixote'),
        factor(c('LOW', 'MED', 'LOW', 'MED', 'MED', 'HIGH'))
        )
```

#### (i) Extract elements $2$ and $1$ of columns $2$ and $3$ of the logical matrix

```{r}
A[[2]][2:1, 2:3]
```

#### (ii) Use `sub` to overwrite `"quixote"` with `"Quixote"` and `"don"` with `"Don"`. Then output to the window:

    "Windmills! ATTACK!"
        -\Don Quixote/-

```{r}
A[[3]][1] <- sub(pattern='d', replacement='D', A[[3]][1])
A[[3]][2] <- sub(pattern='q', replacement='Q', A[[3]][2])

cat(
    '"Windmills! ATTACK"\n\t-\\', 
    A[[3]][1], 
    ' ', 
    A[[3]][2], 
    '/-', 
    sep=''
    )
```

#### (iii) Obtain all values from the sequence between $-4$ and $4$ that are greater than $1$

```{r}
A[[1]][A[[1]] > 1]
```

#### (iv) Using `which`, determine which indexes in the factor are assigned to `"MED"`

```{r}
which(A[[4]] == 'MED')
```

### (b) Create a new list with:

-   The factor from (a) as a component named `"facs"`

-   The numeric vector `c(3, 2.1, 3.3, 4, 1.5, 4.9)` named `"nums"`

-   A nested list made of the first three members of (a) named `"oldlist"`

```{r}
B <- list(
        facs=A[[4]],
        nums=c(3, 2.1, 3.3, 4, 1.5, 4.9),
        oldlist=A[1:3]
        )
```

#### (i) Extract the elements of `"facs"` that correspond to elements of `"nums"` greater than or equal to $3$ 

```{r}
B$facs[B$nums >= 3]
```

#### (ii) Add a new member called `"flags"`, a logical vector of length 6, from a twofold repetition of the third column from the matrix in `"oldlist"`

```{r}
B$flags <- rep(B$oldlist[[2]][, 3], times=2)
```

#### (iii) Use `"flags"` and logical negation to extract entries of `"nums"` corresponding to `FALSE`

```{r}
B$nums[!B$flags]
```

#### (iv) Overwrite the char string in `"oldlist"` with the single string of `"Don Quixote"`

```{r}
B$oldlist[[3]] <- 'Don Quixote'
```

# 5.2 Data Frames

## 5.2.1 Construction

Use `data.frame()`, to create tabular data

```{r}
my_data <- data.frame(
                    person=c(
                            "Peter", 
                            "Lois", 
                            "Meg", 
                            "Chris", 
                            "Stewie"
                            ),
                    age=c(42, 40, 17, 14, 1),
                    sex=factor(c("M", "F", "F", "M", "M"))
                    )
my_data
```

Extract data like matrices

```{r}
my_data[2, 2]
```

```{r}
my_data[3:5, 3]
```

```{r}
my_data[, c(3, 1)]
```

Can access elements with member selection as well

```{r}
my_data$age
```

```{r}
my_data$age[2]
```

Can get dimensions using `nrow(), ncol(), dim()`

```{r}
nrow(my_data)
```

```{r}
ncol(my_data)
```

```{r}
dim(my_data)
```

Character vectors used to be defaulted to factors

```{r}
my_data$person
```

Would need to turn of with `stringsAsFactors=FALSE`

```{r}
my_data <- data.frame(
                    person=c(
                            "Peter", 
                            "Lois", 
                            "Meg", 
                            "Chris", 
                            "Stewie"
                            ),
                    age=c(42, 40, 17, 14, 1),
                    sex=factor(c("M", "F", "F", "M", "M")),
                    stringsAsFactors=FALSE
                    )
my_data
```

```{r}
my_data$person
```

## 5.2.2 Adding Data Columns and Combining Data Frames

Can add new records with `rbind()` , column names and types need to match to properly be added

```{r}
new_record <- data.frame(
                        person="Brian", 
                        age=7, 
                        sex=factor("M", levels=levels(my_data$sex))
                        )
new_record
```

```{r}
my_data <- rbind(my_data, new_record)
my_data
```

Can add columns using `cbind()`

```{r}
funny <- c("High", "High", "Low", "Med", "High", "Med")
funny <- factor(x=funny, levels=c("Low", "Med", "High"))
funny
```

```{r}
my_data <- cbind(my_data, funny)
my_data
```

Can also use member assignment for columns

```{r}
my_data$age.mon <- my_data$age * 12
my_data
```

## 5.2.3 Logical Record Subsets

Can use logical vectors to subset data frames

```{r}
my_data$sex == "M"
```

```{r}
my_data[my_data$sex == "M", ]
```

Can also omit values with negative indices

```{r}
my_data[my_data$sex == "M", -3]
```

Or can use string vectors to select columns

```{r}
my_data[my_data$sex == "M", c("person", "age", "funny", "age.mon")]
```

```{r}
my_data[my_data$age > 10 | my_data$funny == "High", ]
```

Can return a zero row subset, if no criteria is met

```{r}
my_data[my_data$age > 45, ]
```

## 5.2 Exercises

### (a) Create and store the data frame `dframe`:

    person      sex     funny
    Stan        M       High
    Francine    F       Med
    Steve       M       Low
    Roger       M       High
    Hayley      F       Med
    Klaus       M       Med

with the same format as `my_data` from $\S 5.2$

```{r}
dframe <- data.frame(
                    person=c(
                            'Stan', 
                            'Francine', 
                            'Steve', 
                            'Roger', 
                            'Hayley', 
                            'Klaus'
                            ),
                    sex=factor(
                            c(
                            'M', 
                            'F', 
                            'M', 
                            'M', 
                            'F', 
                            'M'
                            )
                            ),
                    funny=factor(
                                c(
                                'High', 
                                'Med', 
                                'Low', 
                                'High', 
                                'Med', 
                                'Med'
                                ),
                                levels=c('Low', 'Med', 'High'),
                                ordered=TRUE
                                )
                    )
```

### (b) Stan and Francine are $41$, Steve is $15$, Hayley is $21$, and Klaus is $60$. Roger is $1,600$ years old. Add this new column

```{r}
dframe$age <- c(41, 41, 15, 1600, 21, 60)
```

### (c) Reorder the columns of `dframe` to be the same as `my_data`

```{r}
dframe <- dframe[c(1, 4, 2, 3)]
```

### (d) Create a new version of `my_data` fro $\S 5.2.2$ called `my_data2` by deleting the `age.mon` column

```{r}
my_data2 <- my_data[, -5]
```

### (e) Combine `my_data2` and `dframe` into one object called `my_data_frame`

```{r}
my_data_frame <- rbind(dframe, my_data2)
```

### (f) Write a single line of code, to extract just the names and ages of any records where the person is female and has a funniness equal to `Med` OR `High`

```{r}
my_data_frame[
            (my_data_frame$sex == 'F') & 
            (my_data_frame$funny != 'Low'), 
            1:2
            ]
```

### (g) Extract all records for those whose name starts with *S*

```{r}
my_data_frame[
    substr(my_data_frame$person, start=1, stop=1) == 'S', ]
```
