---
title: "Chapter 3: Matricies and Arrays"
output: html_notebook
---

# 3.1 Defining a Matrix

Use `matrix()` to define matrices

```{r}
A <- matrix(data=c(-3, 2, 893, 0.17), nrow=2, ncol=2)
A
```

## 3.1.1 Filling Direction

Can control how the matrix is filled with the `byrow` keyword

```{r}
matrix(data=c(1, 2, 3, 4, 5, 6), nrow=2, ncol=3, byrow=FALSE)
```

```{r}
matrix(data=c(1, 2, 3, 4, 5, 6), nrow=2, ncol=3, byrow=TRUE)
```

## 3.1.2 Row and Column Bindings

Can quickly build a matrix from vectors of equal length

Either row-wise

```{r}
rbind(1:3, 4:6)
```

Or column-wise

```{r}
cbind(c(1, 4), c(2, 5), c(3, 6))
```

## 3.1.3 Matrix Dimensions

`dim()` provides the matrix dimensions

```{r}
my_mat <- rbind(c(1, 3, 4), 5:3, c(100, 20, 90), 11:13)
my_mat
```

```{r}
dim(my_mat)
```

`nrow()` and `ncol()` return the first and second values of `dim()` respectively

```{r}
nrow(my_mat)
```

```{r}
ncol(my_mat)
```

```{r}
dim(my_mat)[2]
```

# 3.2 Sub-setting

```{r}
A <- matrix(
    c(0.3, 4.5, 55.3, 91, 0.1, 105.5, -4.2, 8.2, 27.9),
    nrow=3, ncol=3
    )
A
```

Subset matrices with `[m,n]` with `m` being the row number and `n` being the column number

```{r}
A[3, 2]
```

## 3.2.1 Row, Column, and Diagonal Extractions

Can omit values to obtain rows or columns

Omitting the first number will return the column values

```{r}
A[, 2]
```

Omitting the second will return the row values

```{r}
A[1,]
```

Can subset with multiple indices like vectors

```{r}
A[2:3,]
```

```{r}
A[, c(3, 1)]
```

```{r}
A[c(3, 1), 2:3]
```

Can get the diagonal values with `diag()`

```{r}
diag(x=A)
```

## 3.2.2 Omitting and Overwriting

Same as vectors remove elements with negative indices

```{r}
A[, -2]
```

```{r}
A[-1, 3:2]
```

```{r}
A[-1, -2]
```

```{r}
A[-1, -c(2, 3)]
```

Can overwrite elements like vectors

```{r}
B <- A
B
```

```{r}
B[2,] <- 1:3
B
```

```{r}
B[c(1, 3), 2] <- 900
B
```

```{r}
B[, 3] <- B[3,]
B
```

Vector recycling is also possible

```{r}
B[c(1, 3), c(1, 3)] <- c(-7, 7)
B
```

Values will be overwritten row-wise, depending on the given indicies

```{r}
B[c(1, 3), 2:1] <- c(65, -65, 88, -88)
B
```

Can use `diag()` to replace diagonal values

```{r}
diag(x=B) <- rep(x=0, times=3)
B
```

## 3.1 Exercises

### (a) Store a $4 \times 2$ matrix filled row-wise with the values $(4.3, 3.1, 8.2, 8.2, 3.2, 0.9, 1.6, 6.5)$

```{r}
A <- matrix(
    c(4.3, 3.1, 8.2, 8.2, 3.2, 0.9, 1.6, 6.5),
    nrow = 4,
    ncol = 2,
    byrow = TRUE
)
A
```

### (b) Confirm the dimensions of (a) are $3 \times 2$ if you remove any row

```{r}
dim(A[-1, ])
```

### (c) Overwrite the second column of (a) with the same column sorted in ascending order

```{r}
A[, 2] <- sort(A[, 2])
A
```

### (d) What does R return when the fourth row and first column are removed from (c)?

```{r}
matrix(A[-4,-1], ncol = 1)
```

### (e) Store the bottom four elements of (c) as a new $2\times2$ matrix

```{r}
E <- A[3:4, ]
E
```

### (f) Overwrite the elements of (c) at positions $(4,2),(1,2),(4,1),(1,1)$ with $-\frac{1}{2}$ of the values of the diagonal of (e)

```{r}
A[4:1, 2:1] <- diag(E) / 2
A
```

# 3.3 Matrix Operations and Algebra

## 3.3.1 Matrix Transpose

Transpose of $m\times n$ matrix $A$ is $A^T$, an $n\times m$ matrix

$$
A=\left[\begin{array}{cc}
2 & 5 & 2 \\
6 & 1 & 4 
\end{array}\right] 
\Rightarrow 
A^T = \left[\begin{array}{cc}
2 & 6 \\ 
5 & 1 \\ 
2 & 4
\end{array}\right]
$$

```{r}
A <- rbind(c(2, 5, 2), c(6, 1, 4))
A
```

```{r}
t(A)
```

```{r}
t(t(A))
```

## 3.3.2 Identity Matrix

Square $m\times m$ with only ones along the diagonal, use `diag()` to make

$$
I_n = \left[\begin{array}{cc}
1 & 0 & \cdots & 0 & 0 \\ 
0 & 1 & \cdots & 0 & 0 \\ 
\vdots & \vdots & \ddots & \vdots & \vdots \\ 
0 & 0 & \cdots & 1 & 0 \\
 0 & 0 & \cdots & 0 & 1
\end{array}\right]
$$

```{r}
A <- diag(x=3)
A
```

## 3.3.3 Scalar Multiple of a Matrix

Element-wise multiplication with scalars

$$
2 \times \left[ \begin{array}{cc} 
2 & 5 & 2 \\ 
6 & 1 & 4
\end{array} \right] = \left[ \begin{array}{cc}
4 & 10 & 4 \\
12 & 2 & 8
\end{array} \right]
$$

```{r}
A <- rbind(c(2, 5, 2), c(6, 1, 4))
a <- 2
a * A
```

## 3.3.4 Matrix Addition and Subtraction

Element-wise for matrices of equal size

$$
\left[ \begin{array}{cc}
2 & 6 \\
5 & 1 \\
2 & 4
\end{array} \right] - 
\left[ \begin{array}{cc}
-2 & 8.1 \\
3 & 8.2 \\
6 & -9.8
\end{array} \right] =
\left[ \begin{array}{cc}
4 & -2.1 \\
2 & -7.2 \\
-4 & 13.8
\end{array} \right]
$$

```{r}
A <- cbind(c(2, 5, 2), c(6, 1, 4))
A
```

```{r}
B <- cbind(c(-2, 3, 6), c(8.1, 8.2, -9.8))
B
```

```{r}
A - B
```

## 3.3.5 Matrix Multiplication

Matrix $A$ of size $m \times n$ and $B$ of size $p \times q$ , to multiply $n = p$ must be true, $A \cdot B$ will have the size $m \times q$

$$
\left[ \begin{array}{cc}
2 & 5 & 2 \\
6 & 1 & 4
\end{array} \right] \cdot
\left[ \begin{array}{cc}
3 & -3 \\
-1 & 1 \\
1 & 5
\end{array} \right]
$$

$$ \ = \left[ \begin{array}{cc} (2 \times 3) + (5 \times -1) + (2 \times 1) & (2 \times -3) + (5 \times 1) + (2 \times 5) \\ (6 \times 3) + (1 \times -1) + (4 \times 1) & (6 \times -3) + (1 \times 1) + (4 \times 5) \end{array} \right] $$

$$  \ = \left[ \begin{array}{cc} 3 & 9 \\ 21 & 3 \end{array} \right] $$

Matrix multiplication is non-commutative $CD \ne DC$

Cannot use `*` for matrices, need to use `%*%`

```{r}
A <- rbind(c(2, 5, 2), c(6, 1, 4))
dim(A)
```

```{r}
B <- cbind(c(3, -1, 1), c(-3, 1, 5))
dim(B)
```

```{r}
A %*% B
```

```{r}
B %*% A
```

## 3.3.6 Matrix Inversion

Some square matrices can be inverted, where $AA^{-1} = I_m$

$$
\left[ \begin{array}{cc}
3 & 1 \\
4 & 2
\end{array} \right]^{-1} =
\left[ \begin{array}{cc}
1 & -0.5 \\
-2 & 1.5
\end{array} \right]
$$

Non-invertible matrices are called *singular*

Use `solve()` to invert matrix

```{r}
A <- matrix(data=c(3, 4, 1, 2), nrow=2, ncol=2)
A
```

```{r}
solve(A)
```

Verify $AA^{-1}=I_m$

```{r}
A %*% solve(A)
```

## 3.2 Exercises

### (a) Calculate:

$$
\frac {2}{7} \left(
\left[ \begin{array}{cc}
1 & 2 \\
2 & 4 \\
7 & 6
\end{array} \right] - 
\left[ \begin{array}{cc}
10 & 20 \\
30 & 40 \\
50 & 60
\end{array} \right]
\right)
$$

```{r}
A_1 <- rbind(c(1, 2), c(2, 4), c(7, 6))
A_2 <- cbind(c(10, 30, 50), c(20, 40, 60))
(2 / 7) * (A_1 - A_2)
```

### (b) Store the following, and calculate if possible:

$$
A = \left[ \begin{array}{cc}
1 \\
2 \\
7
\end{array} \right], \ \ \
B = \left[ \begin{array}{cc}
3 \\
4 \\
8
\end{array} \right]
$$

```{r}
B_1 <- matrix(c(1, 2, 7), nrow=3, ncol=1)
B_2 <- matrix(c(3, 4, 8), nrow=3, ncol=1)
```

#### (i) $A \cdot B$

```{r}
B_1 %*% B_2 # not possible
```

#### (ii) $A^T \cdot B$

```{r}
t(B_1) %*% B_2
```

#### (iii) $B^T \cdot (A \cdot A^T)$

```{r}
t(B_2) %*% (B_1 %*% t(B_1))
```

#### (iv) $(A \cdot A^T) \cdot B^T$

```{r}
(B_1 %*% t(B_1)) %*% t(B_2) # not possible
```

#### (v) $\left[ \left(B \cdot B^T \right) + \left(A \cdot A^T \right) - 100I_3 \right]^{-1}$

```{r}
solve(
    (B_2 %*% t(B_2)) + (B_1 %*% t(B_1) - (100 * diag(3)))
)
```

### (c) Confirm $A^{-1} \cdot A -I_4$ is a $4 \times 4$ matrix of zeros, where:

$$
A = \left[ \begin{array}{cc}
2 & 0 & 0 & 0 \\
0 & 3 & 0 & 0 \\
0 & 0 & 5 & 0 \\
0 & 0 & 0 & -1
\end{array} \right]
$$

```{r}
C <- diag(c(2, 3, 5, -1))
(solve(C) %*% C) - diag(4)
```

# 3.4 Multidimensional Arrays

1D -\> Vector, 2D -\> Matrix, \>3D -\> Arrays

## 3.4.1 Definition

Use the `array()` function, and specify the individual elements as a vector, as well as the size as another vector

3D Array:

```{r}
AR <- array(data=1:24, dim=c(3, 4, 2))
AR
```

4D Array:

```{r}
BR <- array(data=rep(1:24, times=3), dim=c(3, 4, 2, 3))
BR
```

## 3.4.2 Subsets, Extraction, and Replacements

Arrays are indexed in the same manor as vectors and arrays

Second row of the second layer

```{r}
AR[2,, 2]
```

Third and first elements of above

```{r}
AR[2, c(3, 1), 2]
```

First rows from both layers, returned as columns by default

```{r}
AR[1,,]
```

Second row, first column, first layer, third block

```{r}
BR[2, 1, 1, 3]
```

First row of the first block

```{r}
BR[1,,, 1]
```

Second layer

```{r}
BR[,, 2,]
```

Extracted values of d-dimensional arrays are returned as d+1 arrays

```{r}
BR[3:2, 4,,]
```

```{r}
BR[2,, 1,]
```

## 3.3 Exercise

### (a) Create a 3D array, with $6$ layers of a $4 \times 2$ matrix of the decreasing sequence of $4.8$ to $0.1$ of appropriate length

```{r}
A <- array(data = seq(4.8, 0.1, length.out = 8),
           dim = c(4, 2, 6))
A
```

### (b) Extract the fourth and first row elements of the second column of all layers of (a)

```{r}
B <- A[c(4, 1), 2,]
B
```

### (c) Use fourfold repetition of the second row of (b) to fill a new array of dimensions $2 \times 2 \times 2 \times 3$

```{r}
C <- array(data = rep(B[2,], times = 4), dim = c(2, 2, 2, 3))
C
```

### (d) Create a new array of deleting the sixth layer of (a)

```{r}
D <- A[, ,-6]
D
```

### (e) Overwrite the second and fourth row elements of the second column of layers $1,3,5$ of (d) with $-99$

```{r}
D[c(2, 4), 2, c(1, 3, 5)] <- -99
D
```
