---
title: "Chapter 8 Reading and Writing Files"
output: html_notebook
---

# 8.1 R-Ready Data Sets

## 8.1.1 Built-in Data Sets

`datasets` library gives a set of built-in datasets available in base-r

Can get a summary of the data sets with:

```{r}
library(help="datasets")
```

Can access like any other data frame variable

```{r}
ChickWeight[1:15, ]
```

## 8.1.2 Contributed Data Sets

Many other data set libraries available, i.e. `tseries` package, first:

`install.packages("tseries")`

```{r}
library("tseries")
```

First have to load data into workspace, then can access

```{r}
data(ice.river)
ice.river[1:5, ]
```

# 8.2 Reading in External Data Files

## 8.2.1 The Table Format

File created in a tabular format, called *`mydatafile.txt`*

Can then load file in using `read.table()`

```{r}
my_data_file <- read.table(
                            file='./data/mydatafile.txt',
                            header=TRUE,
                            sep=" ",
                            na.strings="*",
                            stringsAsFactors=FALSE
                        )
my_data_file
```

`file` keyword will use linux style file paths, can use `list.files()` like `ls`

```{r}
list.files("./data")
```

`file.choose()` will open a file explorer, and return the correctly formatted file path string

```{r}
file.choose()
```

Can then pass to `read.table()`, and work with the data

```{r}
my_data_file$sex <- as.factor(my_data_file$sex)
my_data_file$funny <- factor(
                            x=my_data_file$funny, 
                            levels=c("Low", "Med", "High")
                            )
```

## 8.2.2 Spreadsheet Workbooks

Created spreadsheet in *`spreadsheetfile.xlsx`* then saved to *`shreadsheetfile.csv`*

Then use `read.csv()` to load data in

```{r}
spread <- read.csv(
                file='./data/spreadsheetfile.csv', 
                header=FALSE, 
                stringsAsFactors=TRUE
                )
spread
```

## 8.2.3 Web-Based Files

Can read in data from URLs with `read.table()`

```{r}
dia.url <- "https://www.amstat.org/publications/jse/v9n2/4Cdata.txt"
diamonds <- read.table(dia.url)
```

```{r}
names(diamonds) <- c("Carat", "Colour", "Clarity", "Cert", "Price")
diamonds[1:5, ]
```

## 8.2.4 Other File Formats

R can work with many other file types, and third party packages can make this even easier

# 8.3 Writing Out Data Files and Plots

## 8.3.1 Data Sets

Write data to file using `write.table()`

```{r}
write.table(
            x=my_data_file,
            file="./output/somenewfile.txt",
            sep="@",
            na="??",
            quote=FALSE,
            row.names=FALSE
            )
```

`write.csv()` is also available, `file.choose()` is not usually used here

## 8.3.2 Plots and Graphics Files

Create *.jpeg* files using the `jpeg()` object graphics device

```{r}
jpeg(
    filename="./output/myjpegplot.jpeg",
    width=600,
    height=600
    )
plot(
    1:5, 6:10, 
    ylab="a nice ylab", 
    xlab="here's an xlab", 
    main="a saved .jpeg plot"
    )
points(1:5, 10:6, cex=2, pch=4, col=2)
dev.off()
```

Need to call `dev.off()` to finish plot and write file

Create *.pdf* with `pdf()` objects

```{r}
pdf(file="./output/mypdfplot.pdf", width=5, height=5)
plot(
    1:5, 6:10,
    ylab="a nice ylab",
    xlab="here's an xlab",
    main="a saved .pdf plot"
    )
points(1:5, 10:6, cex=2, pch=4, col=2)
dev.off()
```

Can use `ggplot2` plots `ggsave()` function

```{r fig.height=8, fig.width=8}
library("ggplot2")

foo <- c(1.1, 2, 3.5, 3.9, 4.2)
bar <- c(2, 2.2, -1.3, 0, 0.2)

(
qplot(foo, bar, geom='blank')
+ geom_point(size=3, shape=8, color="darkgreen")
+ geom_line(color="orange", linetype=4)
)
```

```{r}
ggsave(filename='./output/mypngplot.png')
```

# 8.4 Ad-Hoc Object Read/Write Operations

Can read/write R-objects to files

```{r}
somelist <- list(
                foo=c(5, 2, 45),
                bar=matrix(
                            data=c(T, T, F, F, F, F, T, F, T), 
                            nrow=3, 
                            ncol=3
                            ),
                baz=factor(
                            c(1, 2, 2, 3, 1, 1, 3),
                            levels=1:3,
                            ordered=TRUE
                        )
                )
somelist
```

Use of `dput()` to write R-objects

```{r}
dput(x=somelist, file='./output/myRobject.txt')
```

Then use `dget()` to retrieve object from file

```{r}
new_object <- dget(file='./output/myRobject.txt')
new_object
```

## 8.1 Exercises

### (a) Using the built-in data set `quakes`:

#### (i) Select only the records where the `mag` is greater than or equal to 5, then write it to a table-format file called *`q5.txt`*, using a delimiter of `!` and now row names

```{r}
q5 <- quakes[quakes$mag >= 5, ]
write.table(
            q5, 
            './output/q5.txt', 
            sep='!', 
            quote=FALSE, 
            row.names=FALSE
            )
```

#### (ii) Read the file back into an object called `q5.dframe`

```{r}
q5.dframe <- read.table(
                        './output/q5.txt', 
                        header=TRUE, 
                        sep='!', 
                        stringsAsFactors=FALSE
                        )
q5.dframe
```

### (b) In the `car` package, use the data frame called `Duncan`. Then:

#### (i) Plot `education` on the x-axis and `income` on the y-axis, with fixed limits of $[0, 100]$, with appropriate labels. For jobs with a `prestige` less than or equal to $80$ use a black $\circ$, for greater than $80$  use a blue $\bullet$  

```{r fig.height=8, fig.width=8}
library(car)

x <- Duncan$education
y <- Duncan$income

plot(
    x, y, 
    type='n', 
    xlim=c(0, 100), 
    ylim=c(0, 100), 
    xlab='Percentage of people with a high school education', 
    ylab='Percentage of people with income >$3.5k / yr', 
    main='Education vs Income for Jobs in 1950'
    )

points(
        x[Duncan$prestige <= 80], 
        y[Duncan$prestige <= 80], 
        col='black', 
        pch=1
        )
points(
        x[Duncan$prestige > 80], 
        y[Duncan$prestige > 80], 
        col='blue', 
        pch=20
        )
```

#### (ii) Add a legend explaining the differences, then save a \$500 \\times 500\$ pixel .png file

```{r}
png('./output/8_1_b.png', width=500, height=500)

plot(
    x, y, 
    type='n', 
    xlim=c(0, 100), 
    ylim=c(0, 100), 
    xlab='Percentage of people with a high school education', 
    ylab='Percentage of people with income >$3.5k / yr', 
    main='Education vs Income for Jobs in 1950'
    )

points(
        x[Duncan$prestige <= 80], 
        y[Duncan$prestige <= 80], 
        col='black', 
        pch=1
        )
points(
        x[Duncan$prestige > 80], 
        y[Duncan$prestige > 80], 
        col='blue', 
        pch=20
        )

legend(
    'topleft',
    legend=c('Job prestige \u2264 80%', 'Job prestige > 80%'),
    pch=c(1, 20),
    col=c('black', 'blue'),
    pt.cex=c(2, 2)
)

dev.off()
```

### (c) Create a list called `exer` that contains the three data sets `quakes`, `q5.dframe`, and `Duncan` Then:

```{r}
exer <- list(quakes=quakes, q5.dframe=q5.dframe, Duncan=Duncan)
```

#### (i) Write the list to disk, called *`Exercise8-1.txt`*

```{r}
dput(exer, './output/Exercise8-1.txt')
```

#### (ii) Read *`Exercise8-1.txt`*, call the result `list.of.dataframes`

```{r}
list.of.dataframes <- dget('./output/Exercise8-1.txt')
names(list.of.dataframes)
```

### (d) Use `ggsave()` to save a copy of the figure from $\S 7.4.3$ as a *.tiff* file

```{r fig.height=8, width=8}
x <- 1:20
y <- c(
        -1.49, 3.37, 2.59, -2.78, -3.94, 
        -0.92, 6.43, 8.51, 3.41, -5.23, 
        -12.01, -6.58, 2.87, 14.12, 9.63, 
        -4.58, -14.75, -11.67, 1.17, 15.62
        )

ptype <- rep(NA, length(x))
ptype[y >= 5] <- 'too_big'
ptype[y <= -5] <-'too_small'
ptype[(x >= 5 & x <= 15) & (y > -5 & y < 5)] <- 'sweet'
ptype[(x < 5 | x > 15) & (y > -5 & y < 5)] <- 'standard'
ptype <- factor(ptype)

(
qplot(x, y, color=ptype, shape=ptype) 
+ geom_point(size=4) 
+ geom_line(mapping=aes(group=1), color='black', lty=2) 
+ geom_hline(mapping=aes(yintercept=c(-5, 5)), color='red')
+ geom_segment(
                mapping=aes(x=5, y=-5, xend=5, yend=5), 
                color='red', 
                lty=3
                ) 
+ geom_segment(
                mapping=aes(x=15, y=-5, xend=15, yend=5), 
                color='red', 
                lty=3
                )
)
```

```{r}
ggsave(
        filename='./output/Section7-4-3.tiff', 
        width=180, 
        height=150, 
        units='mm'
        )
```
