---
title: "Chapter 9 Calling Functions"
output: html_notebook
---

# 9.1 Scoping

## 9.1.1 Environments

### Global Environment

All objects created in current session, can view global objects with `ls()`

```{r}
foo <- 4 + 5
bar <- "stringtastic"
ls()
```

### Package Environments and Namespace

Items made available via packages and `library()`

```{r}
ls("package:graphics")
```

### Local Environments

Objects passed to functions as parameters / arguments, i.e. `data` vector:

```{r}
youthspeak <- matrix(
                    data=c("OMG", "LOL", "WTF", "YOLO"), 
                    nrow=2, 
                    ncol=2
                    )
youthspeak
```

## 9.1.2 Search Path

Can view the R search path with `search()`

```{r}
search()
```

R will search the path from `.GlobalEnv` to `package:base` to find an object

Can use `environment()` to find where the object is located

```{r}
environment(seq)
```

```{r}
environment(arrows)
```

Loading a library with insert the package in the search path

```{r}
library("car")
```

```{r}
search()
```

An error will occur when an object is not found

```{r}
neither.here()
```

```{r}
nor.there
```

## 9.1.3 Reserved and Protected Names

Some words are reserved and forbidden to be used as an object name

-   `if` and `else`

-   `for`, `while`, and `in`

-   `function`

-   `repeat`, `break`, and `next`

-   `TRUE` and `FALSE`

-   `Inf` and `-Inf`

-   `NA`, `NaN`, and `NULL`

Doing so will result in an error, using a case variant will cause confusion

```{r}
NaN <- 5
```

`TRUE` and `FALSE` are reserved, but the shorthands aren't, be careful

```{r}
T <- 42
F <- TRUE
F&&TRUE
```

Can use `rm()` to remove items from the global environment

```{r}
ls()
```

```{r}
rm(list=ls())
ls()
```

## 9.1 Exercises

### (a) Identify the first $20$ items contained in the built-in and automatically loaded `methods` package. How many are there in total?

```{r}
ls("package:methods")[1:20]
```

```{r}
length(ls("package:methods"))
```

### (b) Determine the environment that owns the following:

#### (i) `read.table`

```{r}
environment(read.table)
```

#### (ii) `data`

```{r}
environment(data)
```

#### (iii) `matrix`

```{r}
environment(matrix)
```

#### (iv) `jpeg`

```{r}
environment(jpeg)
```

### (c) Use `ls()` and a test for character string equality to confirm the function `smoothScatter` is part of the `graphics` package

```{r}
any("smoothScatter" == ls("package:graphics"))
```

# 9.2 Argument Matching

Many ways to use function arguments

## 9.2.1 Exact

Match function argument name exactly, order of arguments doesn't matter

```{r}
bar <- matrix(
            data=1:9,
            nrow=3,
            ncol=3,
            dimnames=list(c("A", "B", "C"), c("D", "E", "F"))
            )
bar
```

```{r}
bar <- matrix(
            nrow=3,
            dimnames=list(c("A", "B", "C"), c("D", "E", "F")),  
            ncol=3,
            data=1:9,            
            )
bar
```

## 9.2.2 Partial

Match part of the argument name, shorter code

```{r}
bar <- matrix(
            nr=3, 
            di=list(c("A", "B", "C"), c("D", "E", "F")),
            nc=3,
            dat=1:9
            )
bar
```

Cannot shorten fully, due to how argument matching works. Have to be distinct -\> harder to remember than just the exact argument

```{r}
bar <- matrix(
            nr=3,
            di=list(c("A", "B", "C"), c("D", "E", "F")),
            nc=3,
            d=1:9
            )
```

## 9.2.3 Positional

Most compact mode, uses positional matching. Arguments need to be in the correct order.

Can find arguments of a function using `args()`

```{r}
args(matrix)
```

```{r}
bar <- matrix(
            1:9, 3, 3, F, 
            list(c("A", "B", "C"), c("D", "E", "F"))
            )
bar
```

Need to reference all of the arguments prior to use later ones

```{r}
bar <- matrix(1:9, 3, 3, list(c("A", "B", "C"), c("D", "E", "F")))
```

## 9.2.4 Mixed

Perfectly valid to mix types of argument matching

```{r}
bar <- matrix(
            1:9, 3, 3,
            dim=list(c("A", "B", "C"), c("D", "E", "F"))
            )
bar
```

## 9.2.5 Dot-Dot-Dot: Use of Ellipses

`…` used to specify a flexible number of unspecified arguments

```{r}
args(data.frame)
```

or for supplementary arguments:

```{r}
args(plot)
```

## 9.2 Exercises

### (a) Use positional matching with `seq` to create a sequence of values between $-4$ an $4$ in steps of $0.2$

```{r}
seq(-4, 4, 0.2)
```

### (b) Identify which style of argument matching is used in the following:

#### (i) `array(8:1, dim=c(2, 2, 2))`

```{r}
args(array)
```

Positional and exact matching

#### (ii) `rep(1:2, 3)`

```{r}
args(rep)
```

Positional

#### (iii) `seq(from=10, to=8, length=5)`

```{r}
args(seq)
```

Exact and partial (`length.out`)

#### (iv) `sort(descreasing=T, x=c(2, 1, 1, 2, 0.3, 3, 1.3))`

```{r}
args(sort)
```

Exact

#### (v) `which(matrix(c(T, F, T, T), 2, 2))`

```{r}
args(matrix)
```

```{r}
args(which)
```

Positional

#### (vi) `which(matrix(c(T, F, T, T), 2, 2), a=T)`

Positional and partial (`arr.ind`)

### (c) Use the function documentation for `plot.default()` to determine which of the following fall under the ellipsis

-   [ ]  `type`

-   [x] `pch`

-   [ ] `xlab`

-   [ ] `ylab`

-   [x] `lwd`

-   [x] `lty`

-   [x] `col`

```{r}
args(plot.default)
```
